Towards Correct Legion

### 2017-08-26 correctnesshpc/paper.tex
Editing pass. Citations complete.

The strongest guarantees of correctness come from the area of formal
specifications and proofs. A specification is a description of how a program
should behave, while proofs are used to show that an implementation conforms to
a specification. \emph{Formal} specifications and proofs are defined as
specifications and proofs that have been embedded in a machine-checked formal
logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system, jackson1994nuprl}. This process of formalizing specifications and
proofs in machine checked logics has been shown to catch common errors in
previously informal proofs \cite{clarke1996formal}, and therefore is now considered
to provide the highest assurance of correctness currently possible \cite{?}.

@@

For this paper, we take a recent programming model, Legion, and formalize its
semantics and type system. We then prove a few supporting lemmas as examples of
the kinds of things one can prove in an HPC settings. Finally, we discuss the
some of the possible paths towards full, formally correct implementation. This
is no small task, as we discuss later, and the ways forward are numerous and
difficult. But if we believe that HPC informs important decisions, we must
rise to the challenge.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{treichler2013language}. While there has been much work extending Legion since then,
the core semantics has stayed relatively stable \cite{slaughter2015regent, deppart}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges}, 
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{bauer2012legion, treichler2014realm}. That
said, there are still challenges --subtle bugs that occasionally show up in the
implementation. It is possible that these kinds of bugs could manifest
themselves in ways that would lead to incorrect conclusions when informing
major decisions, something we'd generally like to avoid. One goal of this paper
is to move towards formally and systematically removing \emph{all} bugs from
the programming model, giving everyone who uses is far more certainty that 
their code is correct.

@@

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave \cite{dagum1998openmp}. Others, like Legion's,
are written by hand in the using standard notation for logical relations, etc.
For this paper, when we refer to a \emph{formal} specification, we are
referring to a specification defined in a formal logic, eg. Coq.

While semantics defined by hand as logical relations are a huge improvement
over natural language specifications, there is still much room for improvement.
\cite{formalspec vs paper} has shown that even hand-written mathematical
specifications and proofs in computer science papers almost always have errors
that can be caught by a formal proof.  This is not to say, of course, that
formal specifications and proofs are infallible.  On the contrary, it is quite
easy to get a specification wrong. That said, often, if a there is a mistake in
a spec , the process of proving it formally will expose that mistake, allowing
for its correction.

@@

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{treichler2013language}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
is available online at \texttt{https://github.com/lanl/formallegion}.

@@

The language is made up of some standard elements, e.g. \texttt{call},
\texttt{id}, and some Legion-specific constructs e.g. \texttt{partition},
\texttt{upr}. Evaluation is similar in nature, with the standard elements
having standard semantics, and the

@@

\begin{figure}
\centering
\begin{BVerbatim}
Lemma resp_clos_sub : ∀ r1 r2 Ω (M:M), 
  (∀ r1 r2, subr r1 r2 ∈ Ω → lu r1 M ⊆ lu r2 M) → 
           subr r1 r2 ∈* Ω → lu r1 M ⊆ lu r2 M.
...

Lemma resp_clos : ∀ M Ω, M ~ Ω ↔ M ~* Ω.
...
\end{BVerbatim}
\caption{Lemmas about region relations}
\label{regionrelproof}
\end{figure}

\subsection{Full Implementation of the Specification}
In addition to proving theorems and lemmas about the semantics of Legion as they
relate to HPC memory models, another goal of a formal specification is to ensure
both the completeness and soundness of the Core Legion model. Much of the
necessary implementation work here is ensuring that the entire grammar of the
language is both type safe and its operational semantics contain transition
rules for all productions in the language.

@@

As mentioned earlier in the paper, one of the keystones for large formally
correct systems is modularity. This modularity has two purposes. On the one
hand, it enables ease of implementation. This has long been recognized 
in the field of software engineering \cite{?}. The second, analagous advantage,
is that in a similar way modularity eases the burden of proofs. By defining 
specifications as abstractions, one can separate concerns of proofs. For example,
by composing a proof that a program's serial semantics comply with a formal
specification with a proof that a parallel language respects those serial 
semantics, we can get a proof that a parallel program respects the original
formal semantics. In HPC, by applying this idea to parallel programming models like
Legion, we can separate the application developer from having to prove 
correctness with respect to one or more increasingly complex, heterogeneous
architechures. Instead, we can provide an intermediate layer, like Legion, that
allows application developers to reason and prove correctness to much 
simpler operational semantics, and then prove that those semantics are retained
even on the complex, heterogeneous architectures of todays supercomputers.

@@

One common theme in formalizing systems is the discovery of issues in the
corresponding informal specification. For example, when formalizing the C
language for the CompCert project, Leroy et. al. found many cases where the
informal specification was simply incomplete \cite{compcert}. Note that this is
seperate notion from undefined behavior. Like unknown unknowns, gaps in a
specification can make it impossible to prove desired properties of a system.
In contrast, undefined behaviour is only a problem when programs exihibit it.

In formalizing Legion, we discovered that while there is an impressive theory
ensuring that the side effects of concurrent evaluations preserve serial
semantics with respect to memory reads and writes, there is no theory for what
expressions can be evaluated concurrently based on value dependencies. For
example, if we define the program
\begin{center}
\begin{BVerbatim}
Definition read_write x y t T r :=
  elet x t (read (id y)) 
    (write (new T r) (id x)).
\end{BVerbatim}
\end{center}
then the semantics will happily order the \texttt{read} \emph{after} the
\texttt{write}, despite the fact that the value being written, \texttt{x},
generally will need to be read before being written. This isn't catastrophic
for the semantics, as the returned value and resulting heap are equivalent, 
as desired, but it limits the ability to reason effectively about concurrency. 
In the actual implementations this is handled both dynamically with futures
and statically with dependency analysis \cite{regent}.

@@

A third issue is that there is no way to dynamically create and destroy
regions. In other words, it the semantics assumes static memory allocation.
Of course, this differs from the actual implementation, and is an important,
though difficult, semantics to formalize. As memory heirarchies get more 
sophisticated, the need for tools for reasoning about memory use and locality
becomes more essential.

A fourth challenge is that in practice, Legion is not a stand-alone language.
Instead, it is embedded in an existing language. For the current implementations, 
those languages are Regent and C/C++ \cite{regent, legion}. There are two 
interesting paths towards addressing this issue. One would be to build a sort 
of abstraction for languages that Legion is embedded in, and prove correctness
with respect to that abstraction. This would likely be difficult to get right, as 
capturing the important semantics of the languages that it would be embedded in
would likely be a significant challenge. A second possibility for addressing this 
issue would be choosing a particular semantics and focusing on that. In particular,
combining the C API with the CompCert verified C work would likely be a
fruitful, though challenging, venture.

@@

Returning to the "P" in "HPC", one important current area of research is formal
reasoning about performance \cite{performance}. One of the challenges for
Legion, and runtime systems in general, is that the scheduler often has a
non-trivial amount of work to do, leading to large overheads. In Legion, the
combination of reasoning about permissions, memory locality, etc., makes this
overhead particularly hard to predict.

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{?} working towards formally verified complexity of code.
Chlipala et. al. have also used techniques like those described here to construct
and verify hardware description, where performance must be reasoned about down
to single cycles \cite{braibant2013formal}.  This is one area where other tasking
systems, like Cilk, have a better story than Legion. Blumofe et al. informally
describe bounds on runtime overheads, which has been realized in practice as
low overheads in the actual runtime \cite{blumofe1995cilk}.

>>>>

The strongest guarantees of correctness come from the area of formal
specifications and proofs. A specification is a description of how a program
should behave, while proofs are used to show that an implementation conforms to
a specification. \emph{Formal} specifications and proofs are defined as
specifications and proofs that have been embedded in a machine-checked formal
logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
jackson1994nuprl}. This process of formalizing specifications and proofs in
machine checked logics has been shown to catch common errors in previously
informal proofs, and therefore is often considered to provide the highest
assurance of correctness currently possible \cite{clarke1996formal,
yang2011finding}.

@@

For this paper, we take a recent programming model, Legion, and formalize its
semantics and type system. We then prove a few supporting lemmas as examples of
the kinds of things one can prove in an HPC settings. Finally, we discuss the
some of the possible paths towards a full, formally correct implementation.
This is no small task, as we discuss later, and the ways forward are numerous
and difficult. But if we believe that HPC informs important decisions, we must
rise to the challenge.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions of memory. It has
the attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{treichler2013language}. While there has been much work extending
Legion since then, the core semantics has stayed relatively stable
\cite{slaughter2015regent, treichler2016dependent}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges},
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{bauer2012legion,
treichler2014realm}. That said, there are still subtle bugs that occasionally
show up in the implementation. One goal of this paper is to move towards
formally and systematically removing \emph{all} bugs from the programming
model, giving everyone who uses is far more certainty that their code is
correct.

@@

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave \cite{dagum1998openmp}. Others, like
Legion's, are written by hand in the using standard notation for logical
relations. For this paper, when we refer to a \emph{formal} specification, we
are referring to a specification defined in a formal logic, eg. Coq.

While semantics defined by hand as logical relations are a huge improvement
over natural language specifications, there is still much room for improvement.
Like type checking a program, type checking a proof exposes many "bugs" in a
proof. This is not to say, of course, that formal specifications and proofs are
infallible. On the contrary, it is quite easy to get a specification wrong.
That said, often, if a there is a mistake in a spec, the process of proving it
formally will expose that mistake, allowing for its correction.

@@

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{treichler2013language}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics} for semantics. All
code is available online at \texttt{https://github.com/lanl/formallegion}.

@@

The language is made up of some standard elements, e.g. \texttt{call},
\texttt{id}, and some Legion-specific constructs e.g. \texttt{partition},
\texttt{upr}. Evaluation is similar in nature, with the standard elements
having standard semantics with the addition that memory reads and writes must
be valid interleaves of subexpression memory operations. In simpler terms, only
orderings of memory operations that preserve the resulting memory structure are
permitted.

@@

\begin{figure}
\centering
\begin{BVerbatim}
Lemma resp_clos_sub : ∀ r1 r2 Ω (M:M), 
  (∀ r1 r2, subr r1 r2 ∈ Ω → lu r1 M ⊆ lu r2 M) → 
           subr r1 r2 ∈* Ω → lu r1 M ⊆ lu r2 M.

Lemma resp_clos : ∀ M Ω, M ~ Ω ↔ M ~* Ω.
\end{BVerbatim}
\caption{Lemmas about region relations}
\label{regionrelproof}
\end{figure}

\subsection{Specification Properties}
In addition to proving theorems and lemmas about the semantics of Legion as
they relate to HPC memory models, another goal of a formal specification is to
ensure both the completeness and soundness of the Core Legion model. Much of
the necessary implementation work here is ensuring that the entire grammar of
the language is both type safe and its operational semantics contain transition
rules for all productions in the language.

@@

As mentioned earlier in the paper, one of the keystones for large formally
correct systems is modularity. This modularity has two purposes. On the one
hand, it enables ease of implementation. The second, analagous advantage, is
that in a similar way modularity eases the burden of proofs. By defining
specifications as abstractions, one can separate concerns of proofs. For
example, by composing a proof that a program's serial semantics comply with a
formal specification with a proof that a parallel language respects those
serial semantics, we can get a proof that a parallel program respects the
original formal semantics. In HPC, by applying this idea to parallel
programming models like Legion, we can separate the application developer from
having to prove correctness with respect to one or more increasingly complex,
heterogeneous architechures. Instead, we can provide an intermediate layer,
like Legion, that allows application developers to reason and prove correctness
to much simpler operational semantics, and then prove that those semantics are
retained even on the complex, heterogeneous architectures of todays
supercomputers.

@@

One common theme in formalizing systems is the discovery of issues in the
corresponding informal specification. For example, when formalizing the C
language for the CompCert project, Leroy et. al. found many cases where the
informal specification was simply incomplete \cite{leroy2012compcert}. Note
that this is seperate notion from undefined behavior. Like unknown unknowns,
gaps in a specification can make it impossible to prove desired properties of a
system.  In contrast, undefined behaviour is only a problem when programs
exihibit it.

In formalizing Legion, we discovered that while there is an impressive theory
ensuring that the side effects of concurrent evaluations preserve serial
semantics with respect to memory reads and writes, there is no theory for what
expressions can be evaluated concurrently based on value dependencies. For
example, if we define the program
\begin{center}
\begin{BVerbatim}
Definition read_write x y t T r :=
  elet x t (read (id y)) 
    (write (new T r) (id x)).
\end{BVerbatim}
\end{center}
then the semantics will happily order the \texttt{read} \emph{after} the
\texttt{write}, despite the fact that the value being written, \texttt{x},
generally will need to be read before being written. This isn't catastrophic
for the semantics, as the returned value and resulting heap are equivalent, 
as desired, but it limits the ability to reason effectively about concurrency. 
In the actual implementations this is handled both dynamically with futures
and statically with dependency analysis \cite{slaughter2015regent}.

@@

A third issue is that there is no way to dynamically create and destroy
regions. In other words, if the semantics assumes static memory allocation.
Of course, this differs from the actual implementation, and is an important,
though difficult, semantics to formalize. As memory heirarchies get more 
sophisticated, the need for tools for reasoning about memory use and locality
becomes more essential.

A fourth challenge is that in practice, Legion is not a stand-alone language.
Instead, it is embedded in an existing language. For the current implementations, 
those languages are Regent and C/C++ \cite{slaughter2015regent}. There are two 
interesting paths towards addressing this issue. One is to build a sort of
abstraction for languages that Legion is embedded in, and prove correctness
with respect to that abstraction. This would likely be difficult to get right,
as capturing the important semantics of the languages that it would be embedded
in would likely be a significant challenge. A second possibility for addressing
this issue would be choosing a particular semantics and focusing on that. In
particular, combining the C API with the CompCert verified C work would likely
be a fruitful, though challenging, venture.

@@

Returning to the "P" in "HPC", one important current area of research is formal
reasoning about performance \cite{mccarthy2017coq, chargueraud2015machine}. One
of the challenges for Legion, and runtime systems in general, is that the
scheduler often has a non-trivial amount of work to do, leading to large
overheads. In Legion, the combination of reasoning about permissions, memory
locality, etc., makes this overhead particularly hard to predict.

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{mccarthy2017coq} working towards formally verified
complexity of code.  Chlipala et. al. have also used techniques like those
described here to construct and verify hardware description, where performance
must be reasoned about down to single cycles \cite{braibant2013formal}.  This
is one area where other tasking systems, like Cilk, have a better story than
Legion. Blumofe et al. informally describe bounds on runtime overheads, which
has been realized in practice as low overheads in the actual runtime
\cite{blumofe1995cilk}.
